Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2024-08-30 manuscript/manuscript.Rmd
Main fixes to manuscript done.

# Set general theme.
plot_theme <- ggplot2::theme_light(base_size=base_size)
plot_theme$plot.title$size <- ggplot2::rel(1.0)
plot_theme$plot.title$hjust <- 0.5
plot_theme$strip.background <- ggplot2::element_blank()
plot_theme$strip.text$colour <- NULL
plot_theme$plot.margin <- grid::unit(c(2, 2, 2, 2), "points")
plot_theme$legend.title

@@

@@

```{r decreased-normality, echo=FALSE, fig.cap=cap, fig.height=2.5, warning=FALSE}
cap <- paste0(
  "Effect of location shift, scale and outliers on estimation of Box-Cox and Yeo-Johnson transformation parameter $\\lambda$. ",
  "$10000$ samples were drawn from a normal distribution: ",
  "$\\mathcal{N}(\\mu, 1)$ for the shift dataset, $\\mathcal{N}(10, \\sigma)$ for the scale dataset and $\\mathcal{N}(0, 1)$ for the outlier dataset. ",
  "Additionally, an outlier with value $d$ was added to the outlier dataset. ",
  "Since samples are drawn from a normal distribution, a transformation parameter of $\\lambda = 1$ is expected. ",
  "However, a large shift in location, a scale that is small compared to the location, or presence of large outliers ",
  "lead to incorrectly estimated transformation parameter values. "
)

@@

```{r weighting-functions, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=1.5}
cap <- paste0(
  "Weighting functions investigated in this study to make power transformations more robust against outliers. ",
  "The step function was parameterised with $\\delta_1 = 0.60$. ",
  "The triangle and tapered cosine functions were both parameterised with $\\delta_1 = 0.30$ and $\\delta_2 = 0.90$.")

@@

Power transformations aim to transform features to a normal distribution. However, this may not always be successful or possible. Deviations from normality can be detected by normality tests, such as the Shapiro-Wilk test [@Shapiro1965-zd]. In practice, normality tests may be too stringent to assess goodness of fit in presence of large sample sizes, outliers, or both. Here we develop an empirical test for central normality. The null hypothesis $H_0$ is that the distribution is centrally normal. The alternative hypothesis $H_1$ is that the distribution is not centrally normal.

@@

```{r shifted-distributions, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=3.5}
cap <- paste0(
  "Invariant power transformation produces transformation parameters that are invariant to location and shape. ",
  "Samples were drawn from normal, right-skewed and left-skewed distributions, respectively, which then underwent a shift $d_{\\text{shift}}$ or multiplication by $d_{\\text{scale}}$. ",
  "Estimates of transformation parameter $\\lambda$ with the conventional power transformations (circle) show strong dependency on the overall location and scale of the distribution, ",
  "whereas estimates with the invariant power transformations (triangle) are constant."
)

@@

```{=latex}
\begin{table}
\begin{center}
\caption{Optimal weighting parameters and corresponding loss for \textbf{location- and scale-invariant Box-Cox} power transformations. $p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values, and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution. The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column. The {optimal} column shows the optimal parameter values. The \textit{loss} column shows the loss achieved by each method, under optimised parameters. Lower loss indicates better robustness against outliers.}
\label{tab:optimal-weighting-parameters-box-cox}
\begin{tabular}{l r r r r r r r}

@@

```{r optimised-weighting-function-parameters, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=3.5}
cap <- paste0(
  "Robustness of power transformations after optimising weighting function parameters. ",
  "The distribution of errors, i.e. the difference between robustly fitted $\\lambda^r$ ",
  "and expected $\\lambda_{0}$ found in the absence of outliers, is shown for 1000 randomly ",
  "parametrised asymmetric generalised normal distributions with up to 10% outliers on either side ",
  "of the distribution. The top and bottom panel show errors for the invariant Box-Cox ",
  "and invariant Yeo-Johnson transformations, respectively. ",
  "In each panel, the error distribution for the non-robust transformation is shown to the left ",
  "for comparison with different weighting methods. ",
  "Note that a square root transformation was used along the $y$-axis. Median errors are shown above each distribution."
)

@@

```{r empirical-central-normality-residual-error, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "Residual error as a function of empirical probability. ",
  "Percentiles of the error are shown for robust, location- and scale-invariant transformations of 10000 randomly drawn asymmetric ",
  "generalised normal distributions, with 10 randomly drawn outliers distributions, ",
  "(up to 10% of samples) each. Larger errors occur for more extreme values at the edges of each distribution. ",
  "For comparison, the 99th percentile of residual errors for transformations of outlier-free distributions are shown (grey dashed line).")

.plot_residuals(
  plot_theme = plot_theme,
  manuscript_dir = manuscript_dir)

@@

```{r empirical-central-normality-type-1-error-rate, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "Type 1 error rate of transformed asymmetric generalised normal distributions as function of ",
  "the test statistic $\\tau_{\\text{ecn}}$ for the central portion $\\kappa$ of the distribution.")

@@

```{r include = FALSE, eval = FALSE}
# This writes the lookup table for the goodness of fit test to the
# power.transform package.
load(file = "./R/sysdata.rda")

@@

```{=latex}
\begin{table}
\begin{center}
\caption{Test statistic for empirical central normality at $\kappa = 0.80$ as a function of Type I error rate.}
\label{tab:empirical-central-normality}
\begin{tabular}{c c c c c c c}

@@

```{r empirical-central-normality-examples, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Bi-modal distributions and empirical goodness of fit test results. ",
  "The feature (black) is a mixture of two identical sample sets (blue and orange) ",
  "drawn from normal distributions that are offset by a distance $d$. ",
  "We use the empirical goodness of fit test to compute p-values for the hypothesis that ",
  "the distribution is centrally normal. As may be observed, with increasing offset $d$ ",
  "the probability that the feature is centrally normal decreases. ",
  "Quantile-quantile plots are drawn below each distribution."
)

@@

Invariant power transformations are intended to yield improved transformations to normality
in the presence of large shifts in location, scales far from unity, or both.
Above, this characteristic was assessed using simulations. Below, this characteristic
will be evaluated using examples from real datasets.

@@

```{r lung-cancer-age, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Quantile-quantile plots for age of patients with lung cancer, expressed in years. ",
  "The left panel shows the quantile-quantile plot for the original data. ",
  "The middle panel shows the quantile-quantile plots for data transformed using ",
  "the original Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation \\cite{Raymaekers2021-kq}. ",
  "The right panels shows the quantile-quantile plots for data transformed using ",
  "the non-robust and robust invariant transformations."
)

@@

```{r penguin-body-mass, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Quantile-quantile plots for penguin body mass, measured in grams. ",
  "The left panel shows the quantile-quantile plot for the original data. ",
  "The middle panel shows the quantile-quantile plots for data transformed using ",
  "the original Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation \\cite{Raymaekers2021-kq}. ",
  "The right panels shows the quantile-quantile plots for data transformed using ",
  "the non-robust and robust invariant transformations."
)

@@

```{r ames-latitude, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Quantile-quantile plots for latitude coordinate in a housing dataset. ",
  "The left panel shows the quantile-quantile plot for the original data. ",
  "The middle panel shows the quantile-quantile plots for data transformed using ",
  "the original Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation \\cite{Raymaekers2021-kq}. ",
  "The right panels shows the quantile-quantile plots for data transformed using ",
  "the non-robust and robust invariant transformations."
)

out <- .plot_ames_latitude(plot_theme = plot_theme)
out$plot
```
Every transformation reduced the residual sum.
The non-robust location- and scale-invariant transformation did not improve over conventional alternatives and yielded a data distribution lacking central normality (empirical central normality test: $p=0.05$).
TranConventional transformations had high values for the $\lambda$ parameter.

@@

```{r top-gear, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Quantile-quantile plots for fuel consumption data in the Top Gear dataset. ",
  "The dataset contains three strong outliers for highly fuel-efficient vehicles.",
  "The left panel shows the quantile-quantile plot for the original data. ",
  "The empirical central normality test indicates that the original data do not deviate significantly from central normality ($p=0.76$).",
  "The middle panel shows the quantile-quantile plots for data transformed using ",
  "the original Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation \\cite{Raymaekers2021-kq}. ",
  "For these, the empirical central normality test yields $p=0.01$, and $p=0.55$, respectively. ",
  "The right panels shows the quantile-quantile plots for data transformed using ",
  "the non-robust and robust invariant transformations. ",
  "For these, the empirical central normality test yields $p=0.03$, and $p=0.86$, respectively. ",
  "Samples with observed quantiles below $-3.0$ or above $3.0$ are indicated by crosses."
)

@@

```{r ischemic-stroke, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Quantile-quantile plots for the maximum arterial wall thickness in the ischemic stroke dataset [@Romano2022-gq]. ",
  "This data contains two strong outliers. ",
  "The left panel shows the quantile-quantile plot for the original data. ",
  "The empirical central normality test indicates that the original data do not deviate significantly from central normality ($p=0.56$).",
  "The middle panel shows the quantile-quantile plots for data transformed using ",
  "the original Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation [@Raymaekers2021-kq]. ",
  "For these, the empirical central normality test yields $p=0.01$, and $p=0.61$, respectively. ",
  "The right panels shows the quantile-quantile plots for data transformed using ",
  "the non-robust and robust location- and scale-invariant transformations. ",
  "For these, the empirical central normality test yields $p=0.02$, and $p=0.10$, respectively. ",
  "Samples with observed quantiles below $-3.0$ or above $3.0$ are indicated by crosses."
)

@@

To estimate the marginal effects of process parameters, including transformation method, we first fit a regression random forest (ranger package [@Wright2017-rf] (version 0.16.0); 2000 trees, minimum node size 2, other hyperparameters default) with process parameters and task difficulty as predictors and normalised rank as response variable.
The estimated marginal effects are shown in Figure \ref{fig:marginal-effect-plot}.
On the scale of normalised ranks ($[0.0, 1.0]$), the overall estimated marginal effect of using a random forest instead of generalised linear model was $0.272$.
The overall marginal effect of using z-standardisation to normalise features was $0.008$.
Transformation methods had the following marginal effects:
$0.003$ for using conventional Yeo-Johnson transformation instead of no transformation;
$-0.007$ for using robust invariant Yeo-Johnson transformation instead of no transformation; 
$-0.009$ for using robust invariant Yeo-Johnson transformation with empirical central normality test instead of no transformation; 
$-0.010$ for using robust invariant Yeo-Johnson transformation instead of conventional Yeo-Johnson transformation;
and $-0.012$ for using using robust invariant Yeo-Johnson transformation with empirical central normality test instead of conventional Yeo-Johnson transformation.

```{r marginal-effect-plot, echo=FALSE, fig.cap=cap, fig.height=7.5, warning=FALSE}
cap <- paste0(
  "Estimated marginal effect of learners, normalisation and transformation methods on ranked model performance scores in 22800 machine learning experiments on 285 datasets. ",
  "The top-left panel shows the marginal effect of learners, i.e. random forests and generalised linear models (GLM). ",
  "Random forests outperform GLM models for all task difficulties. ",
  "The top-right panel shows the marginal effect of feature normalisation methods, i.e. no normalisation and z-standardisation. ",
  "z-standardisation is generally beneficial, but the estimated effect is neglible. ",
  "The bottom panel shows the marginal effects of different transformation methods, split by learner and normalisation method. ",
  "There is no consistent behaviour, and estimated effects are neglible. ",
  "Note that the ranges of the $x$-axes of the three main panels differ. ",
  "ECTN: empirical central normality test."
)

@@

Robust location- and scale-invariant transformations form a suitable replacement for their conventional counterparts.
They demonstrated robustness against outliers and prevent inaccurate transformations and potential numerical issues due to location and scale of the data distribution.
This is particularly relevant for automated data processing, where such issues may go unnoticed.
However, we did not find a measurable benefit for -- nor detriment to -- model performance in a machine learning experiment of 232 real-world datasets that contained at least one numeric feature.
One reason may be that numeric features with large location shifts ($|\mu| > 1000.0$) were uncommon.
Of the 4886 numeric features in the 232 datasets, 266 (5%) features in 34 datasets had large location shifts, of which 200 appeared in just 2 datasets.

Robust location- and scale-invariant transformations are realised by simultaneously optimising three parameters, i.e. transformation parameter $\lambda$, shift parameter $x_0$ and scale parameter $s$, using maximum likelihood optimisation.
Alternatively, standardisation of a numeric variable (e.g., through subtracting its median value and division by its interquartile range) prior to conventional power transformations may achieve a similar effect in reducing sensitivity to the distribution's location and scale.

This study had the following limitations.
First, in our experimental results we focussed on the Yeo-Johnson transformation because of its ability to handle variables with negative values, and did not consider the Box-Cox transformation.
Likewise, we focussed on maximum likelihood estimation to optimise parameters, and did not investigate other metrics to the same extent, due to lack of demonstrable benefit (Appendix B).

@@

Robust location- and scale-invariant power transformations allow for automated power transformation of numerical features.

>>>>

# Set general theme.
plot_theme <- ggplot2::theme_light(base_size=base_size)
plot_theme$plot.title$size <- ggplot2::rel(1.0)
plot_theme$plot.title$hjust <- 0.5
plot_theme$strip.background <- ggplot2::element_blank()
plot_theme$strip.text$colour <- NULL
plot_theme$plot.margin <- grid::unit(c(2, 2, 2, 2), "points")
plot_theme$legend.key.size <- grid::unit(0.3, 'lines')

@@

To-do

@@

```{r decreased-normality, echo=FALSE, fig.cap=cap, fig.height=2.5, warning=FALSE}
cap <- paste0(
  "Effect of location, scale and outliers on estimation of the Box-Cox and Yeo-Johnson transformation parameter $\\lambda$. ",
  "$10000$ samples were drawn from a normal distribution: ",
  "$\\mathcal{N}(\\mu, 1)$ for the shift dataset, $\\mathcal{N}(10, \\sigma)$ for the scale dataset and $\\mathcal{N}(0, 1)$ for the outlier dataset. ",
  "Additionally, an outlier with value $d$ was added to the outlier dataset. ",
  "Since samples are drawn from a normal distribution, a transformation parameter of $\\lambda = 1$ is expected. ",
  "However, a large shift in location, a scale that is small compared to the location, or presence of large outliers ",
  "lead to incorrectly estimated transformation parameter values."
)

@@

```{r weighting-functions, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=1.5}
cap <- paste0(
  "Weighting functions investigated in this study to make power transformations more robust against outliers. ",
  "In this example, the step function was parameterised with $\\delta_1 = 0.60$. ",
  "The triangle and tapered cosine functions were both parameterised with $\\delta_1 = 0.30$ and $\\delta_2 = 0.90$."
)

@@

Power transformations aim to transform features to a normal distribution.
However, this may not always be successful or possible.
Deviations from normality can be detected by normality tests, such as the Shapiro-Wilk test [@Shapiro1965-zd].
In practice, normality tests may be too stringent with large sample sizes, outliers, or both.
Here we develop an empirical test for central normality.
The null hypothesis $H_0$ is that the distribution is centrally normal.
The alternative hypothesis $H_1$ is that the distribution is not centrally normal.

@@

```{r shifted-distributions, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=3.5}
cap <- paste0(
  "Invariant power transformation produces transformation parameters that are invariant to location and scale. ",
  "Samples were drawn from normal, right-skewed and left-skewed distributions, respectively, which then underwent a shift $d_{\\text{shift}}$ or multiplication by $d_{\\text{scale}}$. ",
  "Estimates of the transformation parameter $\\lambda$ for the conventional power transformations show strong dependency on the overall location and scale of the distribution, ",
  "whereas estimates obtained for the location- and scale-invariant power transformations are constant."
)

@@

```{=latex}
\begin{table}
\begin{center}
\caption{Optimal weighting parameters and corresponding loss for location- and scale-invariant Box-Cox power transformations. $p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values, and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution. The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column. The {optimal} column shows the optimal parameter values. The \textit{loss} column shows the loss achieved by each method, under optimised parameters. Lower loss indicates better robustness against outliers.}
\label{tab:optimal-weighting-parameters-box-cox}
\begin{tabular}{l r r r r r r r}

@@

```{r optimised-weighting-function-parameters, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=3.5}
cap <- paste0(
  "Robustness of power transformations after optimising weighting function parameters. ",
  "The distribution of errors, i.e. the difference between robustly fitted $\\lambda^r$ ",
  "and expected $\\lambda_{0}$ found in the absence of outliers, is shown for 1000 randomly ",
  "parametrised asymmetric generalised normal distributions with up to 10% outliers on either side ",
  "of the distribution. The top and bottom panel show errors for the location- and scale-invariant Box-Cox ",
  "and Yeo-Johnson transformations, respectively. ",
  "In each panel, the error distribution for the non-robust transformation is shown to the left ",
  "for comparison with different weighting methods. ",
  "Note that a square root transformation was applied along the $y$-axis for display purposes. ",
  "Median errors are shown above each distribution."
)

@@

```{r empirical-central-normality-residual-error, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=3.0}
cap <- paste0(
  "Residual error as a function of empirical probability. ",
  "Percentiles of the error are shown for robust, location- and scale-invariant transformations of 10000 randomly drawn asymmetric ",
  "generalised normal distributions, for each of which outliers were randomly drawn 10 times ",
  "(up to 10% of samples). Larger errors occur at the edges of each distribution."
)

.plot_residuals(
  plot_theme = plot_theme,
  manuscript_dir = manuscript_dir
)

@@

```{r empirical-central-normality-type-1-error-rate, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=3.0}
cap <- paste0(
  "Type 1 error rate of transformed asymmetric generalised normal distributions as function of ",
  "the test statistic $\\tau_{\\text{ecn}}$ for the central portion $\\kappa$ of the distribution.")

@@

```{r include = FALSE, eval = FALSE}
# This writes the lookup table for the empirical central normality test to the
# power.transform package.
load(file = "./R/sysdata.rda")

@@

```{=latex}
\begin{table}
\begin{center}
\caption{Test statistic $\tau_{\text{ecn}}$ for empirical central normality at $\kappa = 0.80$ as a function of Type I error rate.}
\label{tab:empirical-central-normality}
\begin{tabular}{c c c c c c c}

@@

```{r empirical-central-normality-examples, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "Bi-modal distributions and empirical central normality test results. ",
  "The feature (black) is a mixture of two identical sample sets (blue and orange) ",
  "drawn from normal distributions that are offset by a distance $d$. ",
  "We use the empirical centrally normality test to compute the probability for the hypothesis that ",
  "the distribution is centrally normal. As may be observed, with increasing offset $d$ ",
  "the probability that the feature is centrally normal decreases. ",
  "Quantile-quantile plots are drawn below each distribution."
)

@@

Location- and scale-invariant power transformations are intended to yield improved transformations to normality in the presence of large shifts in location, distributions that due to location and scale are not centered near zero, or both.
Earlier, we assessed these transformations using simulated data.
In the following, they are evaluated using examples from real datasets.
We focus on the Yeo-Johnson transformations because

@@

```{r lung-cancer-age, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=1.5}
cap <- paste0(
  "Quantile-quantile plots for age of patients with lung cancer, expressed in years. ",
  "The left panel shows the quantile-quantile plot for the original feature ",
  "The middle panel shows the quantile-quantile plots for the feature transformed using ",
  "the conventional Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation. ",
  "The right panel shows the quantile-quantile plots for the feature transformed using ",
  "the non-robust and robust location- and-scale invariant transformations."
)

@@

```{r penguin-body-mass, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=1.5}
cap <- paste0(
  "Quantile-quantile plots for penguin body mass, measured in grams. ",
  "The left panel shows the quantile-quantile plot for the original feature ",
  "The middle panel shows the quantile-quantile plots for the feature transformed using ",
  "the conventional Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation. ",
  "The right panel shows the quantile-quantile plots for the feature transformed using ",
  "the non-robust and robust location- and-scale invariant transformations."
)

@@

```{r ames-latitude, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=1.5}
cap <- paste0(
  "Quantile-quantile plots for latitude coordinate in a housing dataset. ",
  "The left panel shows the quantile-quantile plot for the original feature ",
  "The middle panel shows the quantile-quantile plots for the feature transformed using ",
  "the conventional Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation. ",
  "The right panel shows the quantile-quantile plots for the feature transformed using ",
  "the non-robust and robust location- and-scale invariant transformations."
)

out <- .plot_ames_latitude(plot_theme = plot_theme)
out$plot
```
Every transformation reduced the residual sum.
The non-robust location- and scale-invariant transformation did not improve over conventional alternatives and yielded a data distribution lacking central normality (empirical central normality test: $p=0.05$).
However, conventional transformations had high values for the $\lambda$ parameter, which could lead to numerical issues.

@@

```{r top-gear, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=1.5}
cap <- paste0(
  "Quantile-quantile plots for fuel consumption data in the Top Gear dataset. ",
  "The feature contains three strong outliers related to highly fuel-efficient vehicles.",
  "The left panel shows the quantile-quantile plot for the original feature. ",
  "The empirical central normality test indicates that the original feature does not deviate significantly from central normality ($p=0.76$).",
  "The middle panel shows the quantile-quantile plots for the feature transformed using ",
  "the conventional Yeo-Johnson transformation and Raymaekers and Rousseeuw's robust adaptation. ",
  "For these, the empirical central normality test yields $p=0.01$, and $p=0.55$, respectively. ",
  "The right panel shows the quantile-quantile plots for the feature transformed using ",
  "the non-robust and robust location- and scale-invariant transformations. ",
  "For these, the empirical central normality test yields $p=0.03$, and $p=0.86$, respectively. ",
  "Samples with observed quantiles below $-3.0$ or above $3.0$ are indicated by crosses."
)

@@

```{r ischemic-stroke, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=1.5}
cap <- paste0(
  "Quantile-quantile plots for the maximum arterial wall thickness in patients with ischemic stroke. ",
  "This feature contains two strong outliers. ",
  "The left panel shows the quantile-quantile plot for the original feature. ",
  "The empirical central normality test indicates that the original feature does not deviate significantly from central normality ($p=0.56$). ",
  "The middle panel shows the quantile-quantile plots for the feature transformed using ",
  "the conventional Yeo-Johnson transformation and Raymaekers and Rousseeuw's robust adaptation. ",
  "For these, the empirical central normality test yields $p=0.01$, and $p=0.61$, respectively. ",
  "The right panel show the quantile-quantile plots for the feature transformed using ",
  "the non-robust and robust location- and scale-invariant transformations. ",
  "For these, the empirical central normality test yields $p=0.02$, and $p=0.10$, respectively. ",
  "Samples with observed quantiles below $-3.0$ or above $3.0$ are indicated by crosses."
)

@@

To estimate the marginal effects of process parameters, including transformation method, we first fit a regression random forest (ranger package [@Wright2017-rf] version 0.16.0): 2000 trees, minimum node size 2, other hyperparameters default) with process parameters and task difficulty as predictors and normalised rank as response variable.
The estimated marginal effects are shown in Figure \ref{fig:marginal-effect-plot}.
On the scale of normalised ranks ($[0.0, 1.0]$), the overall estimated marginal effect of using a random forest instead of generalised linear model was $0.272$.
The overall marginal effect of using z-standardisation to normalise features was $0.008$.
Transformation methods had the following marginal effects:
$0.003$ for using conventional Yeo-Johnson transformation instead of no transformation;
$-0.007$ for using robust invariant Yeo-Johnson transformation instead of no transformation; 
$-0.009$ for using robust invariant Yeo-Johnson transformation with empirical central normality test instead of no transformation; 
$-0.010$ for using robust invariant Yeo-Johnson transformation instead of conventional Yeo-Johnson transformation;
and $-0.012$ for using using robust invariant Yeo-Johnson transformation with empirical central normality test instead of conventional Yeo-Johnson transformation.

```{r marginal-effect-plot, echo=FALSE, fig.cap=cap, fig.height=7.5, warning=FALSE}
cap <- paste0(
  "Estimated marginal effect of learners, normalisation and transformation methods on ranked model performance scores in 18560 machine learning experiments on 232 datasets. ",
  "The top-left panel shows the marginal effect of learners, i.e. random forests and generalised linear models (GLM). ",
  "Random forests outperform GLM models for all task difficulties. ",
  "The top-right panel shows the marginal effect of feature normalisation methods, i.e. no normalisation and z-standardisation. ",
  "z-standardisation is generally beneficial, but the estimated effect is neglible. ",
  "The bottom panel shows the marginal effects of different transformation methods, split by learner and normalisation method. ",
  "There is no consistent behaviour, and estimated effects are neglible. ",
  "Note that the ranges of the $x$-axes of the three main panels differ. ",
  "ECNT: empirical central normality test."
)

@@

Robust location- and scale-invariant transformations are a suitable replacement for their conventional counterparts.
They demonstrated robustness against outliers and prevent inaccurate transformations and potential numerical issues due to location and scale of the distribution of a feature.
This is particularly relevant for automated data processing, where such issues may go unnoticed.
However, we did not find a meaningful benefit for -- nor detriment to -- model performance in a machine learning experiment of 232 real-world datasets that contained at least one numeric feature.
One reason may be that numeric features with large location shifts ($|\mu| > 1000.0$) were uncommon.
Of the 4886 numeric features in the 232 datasets, 266 (5%) features in 34 datasets had large location shifts, of which 200 appeared in just 2 datasets.

Location- and scale-invariant transformations are realised by simultaneously optimising three parameters, i.e. transformation parameter $\lambda$, shift parameter $x_0$ and scale parameter $s$.
Alternatively, standardisation of a numeric variable (e.g., through subtracting its median value and division by its interquartile range) prior to conventional power transformations may achieve a similar effect in reducing sensitivity to the distribution's location and scale.

This study had the following limitations.
First, in our experimental results we focussed on the Yeo-Johnson transformation because of its ability to handle features with negative values, and did not evaluate the Box-Cox transformation.
Likewise, we focussed on MLE to optimise parameters, and did not investigate other metrics to the same extent, due to lack of demonstrable benefit (Appendix B).

@@

Compared to their conventional versions, robust location- and scale-invariant Box-Cox and Yeo-Johnson transformations reduce sensitivity to outliers and the location and scale of features.
An empirical central normality test can assess the quality of transformation of features to normal distributions.
The combination of both facilitate the use of power transformations in automated data analysis workflows.
